### 2015-05-20 Harris_2015/manuscript-materials/draft.Rmd
more manuscript

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur against the frequency that would
be expected if *all* species on the landscape were independent
[@strong_ecological_1984; @gotelli_empirical_2009]. Examining a species pair
against such a "null" background, however, rules out the possibility that the
overall association between two species could be driven by an outside force,
as shown in Figure 1. Even though the two shrub species in this example compete 
with one another for resources at a mechanistic level, they end up clustering
together on the landscape because they both grow best in areas that are not overshadowed
by trees. Due to fundamental constraints on the structure of valid
covariance matrices [@brown_constraints_2010], this kind of positive correlation
is almost inevitable, even when all the interactions are negative. For this
reason, knowing that the two shrubs co-occur more than expected by chance
under some null model (e.g. that their correlation in the data set is positive)
can not---by itself---tell us much about their direct effects on one another.

While the competition between the two shrubs in the previous example does not
leave the commonly-expected pattern in community structure (negative
association at the landscape level), it nevertheless does leave a signal in
the data (Figure 1C). Specifically, *among shaded sites*, there will be a
deficit of co-occurrences, and *among unshaded sites*, there will also be such
a deficit. These *conditional* ("all-else-equal") relationships can be estimated with partial
correlations [@albrecht_spatial_2001], although this approach is best-suited
for Gaussian variables.  For non-Gaussian data, a more general approach is
needed.

In this paper, I introduce Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov networks,
which generalize partial correlations to non-Gaussian data, have been used in
many scientific fields to model associations between various kinds of
"particles". For example, a well-studied network called the Ising model has played
an important role in our understanding of physics [where nearby particles tend
to align magnetically with one another; @cipra_introduction_1987]. In spatial
contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers determine which neurons are
connected to one another by modeling the structure in their firing patterns
[@schneidman_weak_2006]. Following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I suggest that ecologists could similarly treat
species as the interacting particles in this modeling framework. Doing so
would allow ecologists to simulate and study the landscape-level consequences
of arbitrary species interaction matrices, even when our observations are not
Gaussian. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously forced
researchers to rely on severe approximations that produced unintelligible
results [e.g. "probabilities" greater than one; @gilpin_factors_1982]. 
Now that it is computationally feasible to fit these models
exactly, the approach has become worth a second look.

@@

***Conditional relationships and Markov networks.***
Ecologists are often interested in inferring direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships. 
The most familiar conditional relationship
is the partial correlation, which indicates the portion of the correlation
between two species that remains after controlling for other variables in the
data set [@albrecht_spatial_2001], as opposed to the simple correlation, which
does not control for other variables. The example with the shrubs and trees in
Figure 1 shows how the two correlation measures can have opposite signs, 
and suggests that the partial correlation is more relevant
for drawing inferences about species interactions (e.g. competition). To the
extent that our observations are not multivariate Gaussian (e.g. 
presence-absence data), partial covariances do not describe variables' conditional
relationships exactly, and a more general approach is needed. Markov networks
allow researchers to generalize conditional relationships to non-Gaussian
networks, much as generalized linear models allow researchers to extend
regression to non-Gaussian response variables [@lee_learning_2012].

@@

A major feature of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of $\beta$
coefficients [@murphy_machine_2012]. For example, if the coefficient linking
two mutualist species is $+2$, then---all else equal---the odds of observing
either species increase by a factor of $e^2$ when its partner is present
[@murphy_machine_2012]. Of course, if all else is *not* equal (e.g. Figure 1,
where the presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from this
expectation. For this reason, it is important to consider how coefficients'
effects propagate through the network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a *partition
function*, $Z(\alpha, \beta)$, which ensures that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). While the analyses in this paper involve exact computations, this
approach quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species the number of bookkeeping operations
required for exact inference quickly spirals exponentially into the billions.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, either through analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but these techniques are beyond the scope of this paper.

@@

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability distribution
defined by the interaction coefficients in that figure. Each of the methods
described below (a Markov network, two correlation-based methods and a null modeling 
approach) was then evaluated on its ability to correctly infer that all
three species interactions were negative, despite the frequent co-occurrences
of the two shrub species.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500, or
2500 observed communities (24 replicate simulations for each combination).
These simulated data sets span the range from small, single-observer data
sets to large collaborative efforts such as the North American Breeding Bird
Survey. As described in Appendix 1, I randomly drew the "true" coefficient
values for each replicate so that most species pairs interacted negligibly, a
few pairs interacted very strongly, and competition was three times more
common than facilitation. I then used Gibbs sampling to randomly generate
landscapes with varying numbers of species and sites via Markov chain Monte
Carlo (Appendix 1). For half of the simulated landscapes, I treated each
species' $\alpha$ coefficient as a constant, as described above. For the other
half, I treated the $\alpha$ coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix 1). The latter set of simulated landscapes provide an
important test of the methods' ability to distinguish co-occurrence patterns
that were generated from pairwise interactions among the observed species from
those that were generated by external forces like abiotic environmental
filtering. This task was made especially difficult because---as with most
analyses of presence-absence data for co-occurrence patterns---the inference
procedure did not have access to any information about the environmental or
spatial variables that helped shape the landscape [cf @connor_checkered_2013;
@blois_framework_2014].

***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
The previous sections involved known values of $\alpha$ and $\beta$. In most
real cases, however, the true values of these parameters are not known, and
ecologists  will need to estimate them from data. When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for
all of the $\alpha$ and $\beta$ coefficients by optimizing Equation 1
(Appendix 2).  Doing so allows us to find the unique set of $\alpha$ and
$\beta$ coefficients that would be expected to produce exactly the observed
occurrence frequencies and co-occurrence frequencies. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum.

>>>>

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur against the frequency that would
be expected if *all* species on the landscape were independent
[@strong_ecological_1984; @gotelli_empirical_2009]. Examining a species pair
against such a "null" background, however, rules out the possibility that the
overall association between two species could be driven by an outside force,
as shown in Figure 1. Even though the two shrub species in this example compete 
with one another for resources at a mechanistic level, they end up clustering
together on the landscape because they both grow best in areas that are not overshadowed
by trees. Thus, even significant deviations from independence cannot---by 
themselves---tell us much about species' direct effects on one another.

While the competition between the two shrubs in the previous example does not
leave the commonly-expected pattern in community structure (negative
association at the landscape level), it nevertheless does leave a signal in
the data (Figure 1C). Specifically, *among shaded sites*, there will be a
deficit of co-occurrences, and *among unshaded sites*, there will also be such
a deficit.

In this paper, I introduce Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them with observational data. Markov networks,
which generalize partial correlations to non-Gaussian data, have been used in
many scientific fields to model associations between various kinds of
"particles". For example, a well-studied network called the Ising model has played
an important role in our understanding of physics [where nearby particles tend
to align magnetically with one another; @cipra_introduction_1987]. In spatial
contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers determine which neurons are
connected to one another by modeling the structure in their firing patterns
[@schneidman_weak_2006]. Following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I suggest that ecologists could similarly treat
species as the interacting particles in this modeling framework. Doing so
would allow ecologists to simulate and study the landscape-level consequences
of arbitrary species interaction matrices, even when our observations are not
Gaussian. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible
results [e.g. "probabilities" greater than one; @gilpin_factors_1982]. 
Now that it is computationally feasible to fit these models
exactly, the approach has become worth a second look.

@@

***Conditional relationships and Markov networks.***
Ecologists are often interested in inferring direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships. 
The most familiar conditional relationship
is the partial correlation, which indicates the portion of the sample correlation
between two species that remains after controlling for other variables in the
data set [@albrecht_spatial_2001]. The example with the shrubs and trees in
Figure 1 shows how the two correlation measures can have opposite signs, 
and suggests that the partial correlation is more relevant
for drawing inferences about species interactions (e.g. competition). 
Markov networks extend this approach to non-Gaussian data, much as
generalized linear models do for linear regression [@lee_learning_2012].

@@

A major benefit of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of $\beta$
coefficients [@murphy_machine_2012]. For example, if the coefficient linking
two mutualist species is $+2$, then---all else equal---the odds of observing
either species increase by a factor of $e^2$ when its partner is present
[@murphy_machine_2012]. Of course, if all else is *not* equal (e.g. Figure 1,
where the presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from this
expectation. For this reason, it is important to consider how coefficients'
effects propagate through the network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a *partition
function*, $Z(\alpha, \beta)$, which ensures that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). Calculating $Z(\alpha, \beta)$ exactly, as is done in this paper,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species the number of bookkeeping operations
required for exact inference quickly spirals exponentially into the billions.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, either through analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but they are beyond the scope of this paper.

@@

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability distribution
defined by the interaction coefficients in that figure (Appendix A). Each of the methods
described below (a Markov network, two correlation-based methods and a null modeling 
approach) was then evaluated on its ability to correctly infer that 
the two shrub species competed with one another, despite their frequent co-occurrence.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500, or
2500 observed communities (24 replicate simulations for each combination; Appendix B).
These simulated data sets span the range from small, single-observer data
sets to large collaborative efforts such as the North American Breeding Bird
Survey. As described in Appendix B, I randomly drew the "true" coefficient
values for each replicate so that most species pairs interacted negligibly, a
few pairs interacted very strongly, and competition was three times more
common than facilitation. I then used Gibbs sampling to randomly generate
landscapes with varying numbers of species and sites via Markov chain Monte
Carlo (Appendix C). For half of the simulated landscapes, I treated each
species' $\alpha$ coefficient as a constant, as described above. For the other
half, I treated the $\alpha$ coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix B). The latter set of simulated landscapes provide an
important test of the methods' ability to distinguish co-occurrence patterns
that were generated from pairwise interactions among the observed species from
those that were generated by external forces like abiotic environmental
filtering. This task was made especially difficult because---as with most
analyses of presence-absence data for co-occurrence patterns---the inference
procedure did not have access to any information about the environmental or
spatial variables that helped shape the landscape [cf @connor_checkered_2013;
@blois_framework_2014].

***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
The previous sections involved known values of $\alpha$ and $\beta$. In practice,
ecologists will often need to estimate these parameters from data instead. 
When the number of species is reasonably small, one can compute exact maximum 
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing 
Equation 1 (Appendix C).  Doing so allows us to find the unique set of $\alpha$ and
$\beta$ coefficients that would be expected to produce exactly the observed
co-occurrence frequencies. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum.

### 2015-05-20 Harris_2015/models.Rmd
more manuscript

@@

scaled_plot = basic_plot +
  scale_shape_manual(values = c(21, 25, 15, 18)) + 
  coord_cartesian(ylim = c(-0.1, 1.01)) + 
  ylab(expression(R^2)) + 
  xlab("Number of sites (log scale)") + 
  scale_x_log10(breaks = c(20, 100, 500, 2500), limits = c(20, 2.5E3)) +
  scale_color_manual(values = c("red3", "purple2", "sienna1", "mediumblue"))

@@

>>>>

# Use the full names of the methods for the plot below
summarized_results$method = factor(
  summarized_results$method, 
  levels = c("rosalia", "partial", "cov", "Null"),
  labels = c("Markov network", "Partial covariance", "Covariance", "Null")
)

# Use the full names of the methods for the plot below
summarized_results$env = factor(
  summarized_results$env, 
  levels = c("FALSE", "TRUE"),
  labels = c("No environment", "Enironment")
)

@@

scaled_plot = basic_plot +
  scale_shape_manual(values = c(21, 25, 15, 18)) + 
  coord_cartesian(ylim = c(-0.12, 1.01)) + 
  ylab(expression(R^2)) + 
  xlab("Number of sites (log scale)") + 
  scale_x_log10(breaks = c(20, 100, 500, 2500), limits = c(20, 2.5E3)) +
  scale_color_manual(values = c("red3", "purple2", "sienna1", "mediumblue"))

@@

## Additional assessment
